Layer dependence as a measure of local dependence

### 2015-02-16 ld7.tex
Changes and revisions

Suppose $u$ and $v$ are percentile ranks of continuous random variables $x$ and $y$.  Then $(u,v)$ has standard uniform marginals and its joint distribution $C$ is a copula \citep{nelson1999ic}.

@@

Layer dependence is intimately connected to the identity
\begin{equation}\label{decompose}
u= \int_0^1 ( u > \alpha )\de\alpha\ .
\end{equation}
 To prove \eref{decompose} note the left hand side is $\int_0^u 1\de \alpha$ equaling the right hand side.  Informally, the derivative of the  ``$\alpha$--layer'' of $u$,  $(\alpha\le u) \de \alpha$ with respect to $u$ is $\infty\de\alpha=1$ when $u=\alpha$ and $0$ otherwise. Thus the $\alpha$--layer captures movements in $u$ solely at $\alpha$, and ignores movements elsewhere. Thus \eref{decompose} shows $u$ is formed from infinitely many $\alpha$--layers, with each layer capturing the variability of $u$ at a different $\alpha$.  Taking expectations of \eref{decompose} yields the familiar result expressing expectation as the integral of the survival function.

Using  \eref{definition} and \eref{decompose}, Spearman's correlation between $u$ and $v$ is
\begin{equation}\label{wtdaverage}
\rho_S\equiv \cor(u,v) =\frac{\cov(u,v)}{\var(u)}=\frac{\int _0^1\cov\{v,(u>\alpha)\}\de\alpha}{1/12}
=\Ex(\ell_\alpha)\; .
\end{equation}
where the expectation $\Ex$ is calculated over $0\le\alpha\le 1$ using density $6\alpha(1-\alpha)$. Hence Spearman's correlation averages $\alpha$--layer dependence using a certain density on $\alpha$. The density integrates to $1$, has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Varying the density leads to different emphasis on different areas of the relationship between $u$ and $v$, and yields alternative measures of overall dependence. For example using the density $n\alpha^{n-1}$ where $n>0$ puts increasing weight on upper tail dependence. This is further explored in \sref{saltoverall}.

@@

If $(u,v)$ is exchangeable,  $C(u,v)=C(v,u)$, then layer dependence $\ell_\alpha$ measures the lack of discordance and dispersion at $\alpha$:
\begin{equation}\label{decomposition}
\ell_\alpha = 1-2(1+\gamma_\alpha) \delta_\alpha \;,
\end{equation}
where
$$
\gamma_\alpha \equiv \cor\{(u\leq\alpha),(v>\alpha)\} =\cor\{(u>\alpha),(v\leq\alpha)\}\;,
$$
$$
\delta_\alpha \equiv\E\left\{(|u-v|)|(u-\alpha)(v-\alpha)<0\right\}  \; .
$$
A proof of \eref{decomposition} is shown below.  The correlation $-1\leq\gamma_\alpha\leq 1$ measures the tendency for $(u,v)$ to be discordant at $\alpha$: opposite signs on $u-\alpha$ and $v-\alpha$.  The expectation $0\leq \delta_\alpha\leq 1$ measures the dispersion between discordant points $u$ and $v$ at $\alpha$.

@@

The following proposes an approach to fit copulas to data using layer dependence. The approach produces a better fit than past data, and easily incorporates expert opinion on the dependence structure. The approach is illustrated using historical NASDAQ and FTSE stock returns.

@@

\item Smooth calculated layer dependence curve either parametrically or semi-parametrically to remove random error in past data.

@@

\item Fit a copula to refined layer dependence curve, for example using the factor copula model described in the next subsection.
\end{enumerate}
The layer dependence approach offers two advantages over fitting parametric copulas. Firstly layer dependence reflects the dependence structure exhibited in past data, whereas parametric copulas, with their relatively restricted dependence structures, may not properly fit past data. Secondly, layer dependence accommodates expert knowledge of local dependence, whereas parametric copulas permit limited changes to the dependence structure once a parametric form is selected.

@@

Suppose $s$ is a random variable. The factor copula model is
\begin{equation}\label{factor}
u=F(s+\eps) \cq
v=F(s+\eta)    \cq \eps, \eta \sim N(0,1) \ ,
\end{equation}
where $s$, $\eps$ and $\eta$ are independent and $F$ is the  distribution of the random variables in the brackets. Then both $u$ and $v$ are uniform and the joint distribution of $(u,v)$ is exchangeable. Exchangeability is assumed since most commonly used copulas, such as Archimedean copulas \citep{mcneil2005qrm}, are exchangeable. The factor copula model in \eref{factor} is made non-exchangeable for example by dropping either $\eps$ and $\eta$.

@@

The following illustrates layer dependence copula fitting by considering NASDAQ and FTSE returns. The top left panel in \fref{fmarketindex} plots percentile ranks of daily NASDAQ and FTSE returns from $1985$ and $2014$. Layer dependence calculated at every integer percentile and Spearman's correlation are shown in the same panel. Spearman's correlation is $0.4$, indicating moderate dependence between NASDAQ and FTSE returns. Layer dependence increases towards both tails, particularly in the lower tail. Hence major corrections in NASDAQ and FTSE tend to occur simultaneously, and negative corrections are more strongly dependent than positive corrections.

The top right panel in \fref{fmarketindex} fits a factor copula to a smooth layer dependence curve passing through empirical values\footnote{The smoothing is performed by fitting a polynomial of order 4 using least squares estimation.}. Simulated values of the fitted factor copula are shown in the same panel. The bottom left panel fits a Gaussian copula to past data by matching Spearman's correlation. The Gaussian copula has a symmetric layer dependence curve, and hence does not capture the stronger lower tail dependence exhibited in NASDAQ and FTSE data. The same inadequacy applies to Student-$t$ and other elliptical copulas.

@@

>>>>

Suppose $u$ and $v$ are percentile ranks of continuous random variables $x$ and $y$.  Then $(u,v)$ has standard uniform marginals and joint distribution $C$, a copula \citep{nelson1999ic}.

@@

Layer dependence relates  to the identity
\begin{equation}\label{decompose}
u= \int_0^1 ( u > \alpha )\de\alpha\ .
\end{equation}
 To prove \eref{decompose} note the left hand side is $\int_0^u 1\de \alpha$ equaling the right hand side.  Informally, the derivative of the  ``$\alpha$--layer'' of $u$,  $(\alpha\le u) \de \alpha$ with respect to $u$ is $\infty\de\alpha=1$ when $u=\alpha$ and $0$ otherwise. Thus the $\alpha$--layer captures movements in $u$ solely at $\alpha$, and ignores movements elsewhere. Thus \eref{decompose} shows $u$ is formed from infinitely many $\alpha$--layers, with each layer capturing the variability of $u$ at a different $\alpha$.  Taking expectations of \eref{decompose} yields the familiar result expressing expectation as the integral of the survival function.

Using  \eref{definition} and \eref{decompose}, Spearman's correlation between $u$ and $v$ is
\begin{equation}\label{wtdaverage}
\rho_S\equiv \cor(u,v) =\frac{\cov(u,v)}{\var(u)}=\frac{\int _0^1\cov\{v,(u>\alpha)\}\de\alpha}{1/12}
=\Ex(\ell_\alpha)\; .
\end{equation}
where the expectation $\Ex$ is calculated over $0\le\alpha\le 1$ using density $6\alpha(1-\alpha)$.  The density integrates to $1$, has minimum $0$ at $\alpha=0$ and $1$, and increases symmetrically to maximum at $\alpha=0.5$. Modifying the density leads to different emphasis on different areas of the relationship between $u$ and $v$, and yields alternative measures of overall dependence. For example using the density $n\alpha^{n-1}$ where $n>0$ puts increasing weight on upper tail dependence. This is further explored in \sref{saltoverall}.

@@

If $(u,v)$ is exchangeable,  $C(u,v)=C(v,u)$, then layer dependence $\ell_\alpha$ measures the lack of discordance and dispersion at $\alpha$:
\begin{equation}\label{decomposition}
\ell_\alpha = 1-2(1+\gamma_\alpha) \delta_\alpha \;,
\end{equation}
where
$$
\gamma_\alpha \equiv \cor\{(u\leq\alpha),(v>\alpha)\} =\cor\{(u>\alpha),(v\leq\alpha)\}\;,
$$
$$
\delta_\alpha \equiv\E\left\{(|u-v|)|(u-\alpha)(v-\alpha)<0\right\}  \; .
$$
A proof of \eref{decomposition} is  below.  The correlation $-1\leq\gamma_\alpha\leq 1$ measures the tendency for $(u,v)$ to be discordant at $\alpha$: opposite signs on $u-\alpha$ and $v-\alpha$.  The expectation $0\leq \delta_\alpha\leq 1$ measures the dispersion between discordant points $u$ and $v$ at $\alpha$.

@@

This section proposes fitting a copula  to data using layer dependence. The approach can  incorporate expert opinion on the dependence structure. The approach is illustrated using historical NASDAQ and FTSE stock returns.

@@

\item Smooth calculated layer dependence curve either parametrically or semi-parametrically.

@@

\item Fit a copula given the fitted layer dependence curve, for example using the factor copula model described in the next subsection.
\end{enumerate}
The layer dependence approach offers two advantages over fitting parametric copulas. Firstly layer dependence reflects the dependence structure exhibited in past data, whereas parametric copulas, with their relatively restricted dependence structures, may not properly fit past data. Secondly, layer dependence accommodates expert knowledge of local dependence, whereas parametric copulas permit limited changes to the dependence structure once a parametric form is selected.

@@

Suppose $s$ is a random variable. The factor copula model is
\begin{equation}\label{factor}
u=F(s+\eps) \cq
v=F(s+\eta)    \cq \eps, \eta \sim N(0,1) \ ,
\end{equation}
where $s$, $\eps$ and $\eta$ are independent and $F$ is the  distribution of the random variables in the brackets. Then both $u$ and $v$ are uniform and the joint distribution of $(u,v)$ is exchangeable. Exchangeability is assumed since most commonly used copulas, such as Archimedean copulas \citep{mcneil2005qrm}, are exchangeable. The factor copula model in \eref{factor} is made non-exchangeable for example by dropping either $\eps$ or $\eta$.

@@

The following illustrates layer dependence copula fitting using NASDAQ and FTSE returns. The top left panel in \fref{fmarketindex} plots percentile ranks of daily NASDAQ and FTSE returns from $1985$ and $2014$. Layer dependence  at every integer percentile and Spearman's correlation are shown in the same panel. Spearman's correlation is $0.4$, indicating moderate dependence between NASDAQ and FTSE returns. Layer dependence increases towards both tails, but more so in the lower tail. Hence major corrections in NASDAQ and FTSE tend to occur simultaneously, and negative corrections are more strongly dependent than positive corrections.

The top right panel in \fref{fmarketindex} fits a factor copula to a smooth layer dependence curve passing through empirical values\footnote{The smoothing is performed by fitting a polynomial of order 4 using least squares estimation.}. Simulated values of the fitted factor copula are shown in the same panel. The bottom left panel fits a Gaussian copula to past data by matching Spearman's correlation. The Gaussian copula has a symmetric layer dependence curve, and hence does not capture the stronger lower tail dependence exhibited in NASDAQ and FTSE data.  Student-$t$ and other elliptical copulas would suffer from the same inadequacy.

@@

\section*{References}
